---
title: "Spatiotemporal Delineation of Extreme Heat/Cold Events"
subtitle: "Extreme Events Detection"
author:
- affiliation: University of Waterloo
  email: pfard@uwaterloo.ca
  name: Pedram Fard
output:
  html_document:
    self_contained: true
    toc: true
    toc_float:
      toc_collapsed: false
    number_sections: false
    smooth_scroll: true
    toc_depth: 4
    theme: default
    highlight: zenburn
    df_print: paged

  pdf_document: default
  word_document: default

editor_options: 
  chunk_output_type: inline
---

# Workspace Setup
```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, 
                     verbose = TRUE)

knitr::opts_chunk$set(results = "hold", 
                      echo = FALSE, 
                      warning=FALSE, 
                      message = FALSE, 
                      include = TRUE)
```


# I/O Setup
```{r set IO parameters, include=FALSE}
library(here)

spatial_projection  = 4269 #NAD83

project_name <- "EHE_ECE_Detection_"

input_dir <- "data"

output_dir <- (here("outputs","/"))

timestamped <- format(Sys.time(), "%m%d_%H%M")

print(Sys.time)
print(timestamped)

timestamped_folder <- paste0(project_name ,timestamped)

dir.create(file.path(here("outputs"), timestamped_folder))

output_path <- paste0(output_dir, timestamped_folder)

print(output_dir)
print(output_path)
```


## Libraries and Functions
```{r libs}
source(here("R", "libraries.R"))

source(here("R", "helper_functions.R"))

writeLines(capture.output(sessionInfo()), 
           paste0(output_path, "/_", 
                  timestamped, "_session_info.txt"))
```


# Load data
## Detect and handle the EDW outlier values
```{r}
noaa_compiled_sf_selected <- readRDS(here(input_dir, 
                                  "noaa_compiled_sf_selected.rds"))[[1]]

multi_stations_df_adjusted <- readRDS(here(input_dir, 
                                 "multi_stations_df_adjusted.rds"))[[1]]


# tag the records that have the out of range values
stations_adjusted <- multi_stations_df_adjusted %>% 
  mutate(., tag = case_when(
      between(wind_speed_rate, 0, 40) |
      (air_temperature_c < 60) |
      (dewpoint_temperature_c) < 60 ~0,
    TRUE ~ 1)   
  ) %>% 
  mutate(date_day = as.Date(startdate, "%a, %d %b %Y")) %>% #make a new variable to record the days
  mutate(date_week = isoweek(startdate))

hist(stations_adjusted$date_week)

stations_adjusted <- stations_adjusted %>% 
  mutate(., wind_speed_rate = case_when(
      (wind_speed_rate >= 0) &
      (wind_speed_rate < 40) ~wind_speed_rate,
      TRUE ~NA_real_
    )) %>% 
  mutate(., air_temperature_c = case_when(
      (air_temperature_c < 60) ~air_temperature_c,
      TRUE ~NA_real_
    ))  %>%
  mutate(., dewpoint_temperature_c = case_when(
      (dewpoint_temperature_c < 80) ~dewpoint_temperature_c,
      TRUE ~NA_real_
    ))
```


### Plot the adjusted variables histogram
```{r}
hist(stations_adjusted$air_temperature_c)
hist(stations_adjusted$dewpoint_temperature_c)
hist(stations_adjusted$wind_speed_rate)
hist(stations_adjusted$tag)
```


### Plot the adjusted variables scatter plot
```{r}
glimpse(stations_adjusted)

ggplot(stations_adjusted) +
  geom_point(aes(x = startdate, 
                 y = air_temperature_c,
                 color = air_temperature_c))

ggplot(stations_adjusted) +
  geom_point(aes(x = startdate, 
                 y = dewpoint_temperature_c,
                 color = dewpoint_temperature_c))

ggplot(stations_adjusted) +
  geom_point(aes(x = startdate, 
                 y = wind_speed_rate,
                 color = wind_speed_rate))

  #scale_color_gradient(low="blue", high="green")
  #scale_color_gradient(low="yellow", high="red")
```


# Compute apparent temperature
```{r}
# https://rdrr.io/cran/weathermetrics/man/dewpoint.to.humidity.html
# https://www.wpc.ncep.noaa.gov/html/heatindex.shtml
stations_adjusted$derived_humidity <- dewpoint.to.humidity(t = stations_adjusted$air_temperature_c,
                                dp = stations_adjusted$dewpoint_temperature_c,
                                temperature.metric = 'celsius')

apparent_temperature <- apparentTemp(tas=stations_adjusted$air_temperature_c,
                   hurs=stations_adjusted$derived_humidity, 
                   wind=stations_adjusted$wind_speed_rate)

stations_adjusted$apparent_temperature <- unlist(apparent_temperature)

stations_adjusted_hourly_apparent <- stations_adjusted %>% 
            mutate(apparent_actual_temp_diff = apparent_temperature - air_temperature_c) %>% 
    arrange(startdate)
                     
glimpse(stations_adjusted_hourly_apparent)
```


# Plot apparent temperature
```{r, eval = FALSE}
ggplot(stations_adjusted_hourly_apparent) +
  geom_point(aes(x = startdate,
                 y = apparent_temperature,
                 color = apparent_temperature))

ggplot(stations_adjusted_hourly_apparent) +
  geom_point(aes(x = startdate,
                 y = apparent_actual_temp_diff,
                 color = apparent_actual_temp_diff)) +
  scale_color_gradient2(
                        low = "blue", 
                        mid = "white",
                        high = "red")
```


# Compute daily means
```{r}
ls(stations_adjusted_hourly_apparent)

stations_adjusted_dt <- data.table(stations_adjusted_hourly_apparent)

stations_daily_means <- stations_adjusted_dt[, 
        .(dm_apparent_temperature = round(mean(apparent_temperature, na.rm = T),2),
          dm_air_temperature_c = round(mean(air_temperature_c, na.rm = T),2),
          dm_sea_level_pressure = round(mean(sea_level_pressure, na.rm = T),2),
          dm_derived_humidity = round(mean(derived_humidity, na.rm = T),2),
          dm_wind_speed_rate = round(mean(wind_speed_rate, na.rm = T),2)
          ), 
        by = .(shape_id, geoid,
               Station_Name = name, 
               latitude, longitude,
               DATE = date_day)]

stations_daily_means_rolling <- stations_daily_means %>% group_by(shape_id) %>% 
  mutate(dm_at_3_0 = round(mean_run(x = dm_apparent_temperature, 
                           k = 3, lag = 0, idx = DATE, 
                           na_rm=TRUE),2)) %>% 
  mutate(dm_at_32_m3 = round(mean_run(x = dm_apparent_temperature, 
                           k = 30, lag = 3, idx = DATE, 
                           na_rm=TRUE),2)) %>% ungroup() %>% 
  #filter(DATE < as.POSIXct("2020-01-01")) %>% 
  data.table()
  
glimpse(stations_daily_means_rolling)
```


# Compute stations percentiles
```{r}
#percentiles
dm_apparent_temperature_percentiles <- stations_daily_means_rolling[ , 
           c(list(median = median(dm_apparent_temperature, na.rm = T)), 
           as.list(round(quantile(dm_apparent_temperature, c(.01, .02, .05, .075, .10, .15, .25,
                                                            .85, .90, .925, .95, .98, .99), 
                                  na.rm = T),2))
            ),
      by = .(shape_id)]

# partial window using adaptive rolling function

dm_apparent_temperature_percentiles

dm_apparent_temperature_compiled <- merge(stations_daily_means_rolling, 
                                  clean_names(dm_apparent_temperature_percentiles), 
                                            by.x="shape_id", 
                                            by.y="shape_id",
                          all.x = TRUE)

glimpse(dm_apparent_temperature_compiled)

ObjSave(stations_adjusted_hourly_apparent,
        stations_daily_means,
        stations_daily_means_rolling,
        dm_apparent_temperature_percentiles,
        dm_apparent_temperature_compiled,
        
        folder = output_path)
```


# Compute Sheridan terms
```{r}
Sheridan <- dm_apparent_temperature_compiled %>%
  # drop those rows with missing apparent temperature (while still can have rolling average)
  #filter(!is.na(dm_apparent_temperature)) %>% 
  
  # set the period of analysis
  filter(DATE < as.POSIXct("2020-01-01")) %>% 
  mutate(EHaccl = dm_at_3_0 - dm_at_32_m3) %>%
  
  mutate(EH = pmax(0, (dm_at_3_0 - x95_percent))) %>% 
  mutate(EHF = pmax(0, EH) * pmax(1, EHaccl)) %>% 
  
  mutate(EC = pmin(0, (dm_at_3_0 - x5_percent))) %>% 
  mutate(ECF = -1 * pmin(0, EC) * pmin(-1, EHaccl))

nrow(dm_apparent_temperature_compiled)
nrow(Sheridan)

hist(Sheridan$EHaccl)

hist(Sheridan$EH)
hist(Sheridan$EHF)

hist(Sheridan$EC)
hist(Sheridan$ECF)
```


# Identify extreme events
```{r}
EHF_percentiles <- Sheridan[ EHF > 0, 
           c(list(median = median(EHF, na.rm = T)), 
           as.list(round(quantile(EHF, c(.01, .02, .05, .10, .15, .25, 
                                         .75, .85, .90, .95, .98, .99), na.rm = T),2))
            ),
      by = .(shape_id)] 

EHF_percentiles
ECF_percentiles <- Sheridan[ ECF < 0, 
           c(list(median = median(ECF, na.rm = T)), 
           as.list(round(quantile(ECF, c(.01, .02, .05, .10, .15, .25, 
                                         .75, .85, .90, .95, .98, .99), na.rm = T),2))
            ),
      by = .(shape_id)]

clean_names(ECF_percentiles)
Sheridan_interim <- merge(Sheridan, 
                           EHF_percentiles %>% setNames(paste0('EHF_', names(.))) %>% clean_names(), 
                                            by.x="shape_id", 
                                            by.y="ehf_shape_id",
                          suffix  = c("", ""),
                          all.x = TRUE)
Sheridan_interim
Sheridan_compiled <- merge(Sheridan_interim, 
                           ECF_percentiles %>% setNames(paste0('ECF_', names(.))) %>% clean_names(), 
                                            by.x="shape_id", 
                                            by.y="ecf_shape_id",
                          suffix  = c("", ""),
                          all.x = TRUE)

hist(EHF_percentiles$`90%`)
hist(EHF_percentiles$`95%`)
hist(EHF_percentiles$`99%`)
# 
hist(ECF_percentiles$`90%`)
hist(ECF_percentiles$`95%`)
hist(ECF_percentiles$`99%`)

ls(Sheridan_compiled)

stations_daily_means_compiled <- Sheridan_compiled %>% 
                        mutate(EHE = case_when(EHF > ehf_85_percent ~ 1,
                               TRUE ~ 0)) %>% 
                        mutate(ECE = case_when(ECF < ecf_15_percent ~ 1,
                               TRUE ~ 0))

stations_daily_means_compiled

stations_daily_means_compiled_sf <- st_as_sf(stations_daily_means_compiled, 
                                   coords = c("longitude", "latitude"),
                                   crs = spatial_projection) %>% as_tibble() %>% st_as_sf()

stations_daily_means_compiled_sp <- as_Spatial(stations_daily_means_compiled_sf, 
                        cast = TRUE)


plot(stations_daily_means_compiled_sf[, "EHF"])
plot(stations_daily_means_compiled_sp[, "EHF"])

glimpse(stations_daily_means_compiled_sf)
glimpse(stations_daily_means_compiled_sp@data)

# makeDataReport(stations_daily_means_compiled, 
#                output = "html",
#                replace = TRUE)
```


## Label consecutive heat/cold days
```{r}
EHE <- stations_daily_means_compiled[, .(EHE = EHE[1], 
                       DATE = DATE, 
                       Month = lubridate::month(DATE),
                       Week = isoweek(DATE),
                       Day = yday(DATE),
                       EHF = EHF,
                       EHMI = EHF * .N,
                       DM_apparent_temperature = dm_apparent_temperature,
                       EHE_day = cumsum(rleidv(EHE)),
                       EHE_duration = .N,
                       EHE_start = first(DATE), 
                       EHE_end = last(DATE),
                       Longitude = longitude,
                       Latitude = latitude,
                       Event_Type = "Extreme Heat Event"
                       ),
      by = .(shape_id, rleidv(EHE))][EHE == 1][, c('EHE', 'rleidv') := NULL][]%>% 
      mutate(UID = paste0(strftime(EHE_start, format = "%Y%m%d"),
                          "1"
                          ))

ECE <- stations_daily_means_compiled[, .(ECE = ECE[1], 
                       DATE = DATE,
                       Month = lubridate::month(DATE),
                       Week = isoweek(DATE),
                       Day = yday(DATE),
                       ECF = ECF,
                       ECMI = ECF * .N,
                       DM_apparent_temperature = dm_apparent_temperature,
                       ECE_day = cumsum(rleidv(ECE)),
                       ECE_duration = .N,
                       ECE_start = first(DATE), 
                       ECE_end = last(DATE),
                       Longitude = longitude,
                       Latitude = latitude, 
                       Event_Type = "Extreme Cold Event"
                       ),
      by = .(shape_id, rleidv(ECE))][ECE == 1][, c('ECE', 'rleidv') := NULL][] %>% 
      mutate(UID = paste0(strftime(ECE_start, format = "%Y%m%d"),
                          "2"
                          ))
```


# Set the cases tudy period (i.e. One Year)
```{r}
# construct table row by row
EHE_ECE <- full_join(EHE, 
                     ECE, 
                                            by.x=c("shape_id", "DATE"), 
                                            by.y=c("shape_id", "DATE"), ,
                          all = TRUE)

# to be used for stats that requires to preserve null values
EHE_ECE_compiled <- merge(stations_daily_means_rolling %>% 
                          filter(DATE < as.POSIXct("2020-01-01")), 
                          
                          EHE_ECE, 
                                            by.x=c("shape_id", "DATE"), 
                                            by.y=c("shape_id", "DATE"), ,
                          all.x = TRUE) %>% 
  mutate_at(vars(Event_Type), ~replace_na(., "Non-Extreme Weather")) 

# To be used for raster creation that requires non-null values
EHE_ECE_compiled_null2zero <- EHE_ECE_compiled %>% 
  mutate_at(vars(ECE_duration, EHE_duration,
                 EHF, EHMI,
                 ECF, ECMI,
                 ), ~replace_na(., 0))

  #https://stackoverflow.com/a/55291983

makeDataReport(EHE_ECE_compiled, 
                output = "html",
                replace = TRUE)

EHE_ECE_compiled_skim_report <- skim(EHE_ECE_compiled) %>% 
     arrange(complete_rate)

EHE_ECE_compiled_skim_report

ls(EHE_ECE_compiled)
```


# Stats
```{r}
EHE_ECE_Aggregate_Annual_Stats <- EHE_ECE_compiled %>% group_by(Event_Type) %>% 
  summarise(Number_of_Records = n(),
            Distinct_Stations = n_distinct(shape_id),
            
            Average_EHE_Duration = round(mean(EHE_duration, na.rm = TRUE), 2),
            Max_EHE_Duration = max(EHE_duration, na.rm = TRUE),
            Average_EHF = round(mean(EHF, na.rm = TRUE), 1),
            Max_EHF = round(max(EHF, na.rm = TRUE), 1),

            Average_EHMI = round(mean(EHMI, na.rm = TRUE), 1),
            Max_EHMI = round(max(EHMI, na.rm = TRUE), 1),

            Average_ECE_Duration = round(mean(ECE_duration, na.rm = TRUE), 2),
            Max_ECE_Duration = max(ECE_duration, na.rm = TRUE),
            Average_ECF = round(mean(ECF, na.rm = TRUE), 1),
            Min_ECF = round(min(ECF, na.rm = TRUE), 1),

            Average_ECMI = round(mean(ECMI, na.rm = TRUE), 1),
            Min_ECMI = round(min(ECMI, na.rm = TRUE), 1),

            Average_AT = round(mean(dm_apparent_temperature, na.rm = TRUE),2),
            Min_AT = min(dm_apparent_temperature, na.rm = TRUE),
            Max_AT = max(dm_apparent_temperature, na.rm = TRUE),
            Median_AT = round(median(dm_apparent_temperature, na.rm = TRUE), 2)
            )
          # Parcels_Total_Area_per_Zone = sum((!!area_field)),
          # Parcels_Mean_Area_per_Zone = mean(!!area_field, na.rm = TRUE)) %>%

EHE_ECE_Aggregate_Monthtly_Stats <- EHE_ECE_compiled %>% group_by(Event_Type, Month) %>% 
  summarise(Number_of_Records = n(),
            Distinct_Stations = n_distinct(shape_id),
            
            Average_EHE_Duration = round(mean(EHE_duration, na.rm = TRUE), 2),
            Max_EHE_Duration = max(EHE_duration, na.rm = TRUE),
            Average_EHF = round(mean(EHF, na.rm = TRUE), 1),
            Max_EHF = round(max(EHF, na.rm = TRUE), 1),

            Average_EHMI = round(mean(EHMI, na.rm = TRUE), 1),
            Max_EHMI = round(max(EHMI, na.rm = TRUE), 1),

            Average_ECE_Duration = round(mean(ECE_duration, na.rm = TRUE), 2),
            Max_ECE_Duration = max(ECE_duration, na.rm = TRUE),
            Average_ECF = round(mean(ECF, na.rm = TRUE), 1),
            Min_ECF = round(min(ECF, na.rm = TRUE), 1),

            Average_ECMI = round(mean(ECMI, na.rm = TRUE), 1),
            Min_ECMI = round(min(ECMI, na.rm = TRUE), 1),

            Average_AT = round(mean(dm_apparent_temperature, na.rm = TRUE),2),
            Min_AT = min(dm_apparent_temperature, na.rm = TRUE),
            Max_AT = max(dm_apparent_temperature, na.rm = TRUE),
            Median_AT = round(median(dm_apparent_temperature, na.rm = TRUE), 2)
            )

EHE_ECE_Stations_Stats <- EHE_ECE_compiled %>% group_by(shape_id, Event_Type) %>% 
  summarise(Number_of_Records = n(),
            Distinct_Events = n_distinct(UID),

            Average_EHE_Duration = round(mean(EHE_duration, na.rm = TRUE), 2),
            Max_EHE_Duration = max(EHE_duration, na.rm = TRUE),
            Average_EHF = round(mean(EHF, na.rm = TRUE), 1),
            Max_EHF = round(max(EHF, na.rm = TRUE), 1),

            Average_EHMI = round(mean(EHMI, na.rm = TRUE), 1),
            Max_EHMI = round(max(EHMI, na.rm = TRUE), 1),

            Average_ECE_Duration = round(mean(ECE_duration, na.rm = TRUE), 2),
            Max_ECE_Duration = max(ECE_duration, na.rm = TRUE),
            Average_ECF = round(mean(ECF, na.rm = TRUE), 1),
            Min_ECF = round(min(ECF, na.rm = TRUE), 1),

            Average_ECMI = round(mean(ECMI, na.rm = TRUE), 1),
            Min_ECMI = round(min(ECMI, na.rm = TRUE), 1),

            Average_AT = round(mean(dm_apparent_temperature, na.rm = TRUE),2),
            Min_AT = min(dm_apparent_temperature, na.rm = TRUE),
            Max_AT = max(dm_apparent_temperature, na.rm = TRUE),
            Median_AT = round(median(dm_apparent_temperature, na.rm = TRUE), 2)
            )

write.csv(EHE_ECE_Aggregate_Annual_Stats, 
          paste0(output_path, "/", 
                 deparse(substitute(EHE_ECE_Aggregate_Annual_Stats)), 
                 "_2019", 
                 ".csv"), 
          row.names = FALSE, )

write.csv(EHE_ECE_Aggregate_Monthtly_Stats, 
          paste0(output_path, "/", 
                 deparse(substitute(EHE_ECE_Aggregate_Monthtly_Stats)), 
                 "_2019", 
                 ".csv"), 
          row.names = FALSE, )

write.csv(EHE_ECE_Stations_Stats, 
          paste0(output_path, "/", 
                 deparse(substitute(EHE_ECE_Stations_Stats)), 
                 "_2019", 
                 ".csv"), 
          row.names = FALSE, )
```


# Export data objects as RDS files
```{r}
ObjSave(EHE,
        ECE,
        Sheridan_compiled,
        EHF_percentiles,
        ECF_percentiles,
        dm_apparent_temperature_percentiles,
        EHE_ECE,
        EHE_ECE_compiled,
        EHE_ECE_compiled_null2zero,
        
        folder = output_path)


write.csv(dm_apparent_temperature_percentiles, 
          paste0(output_path, "/", 
                 deparse(substitute(dm_apparent_temperature_percentiles)), 
                 "_2019_", 
                 nrow(dm_apparent_temperature_percentiles), 
                 "stations", 
                 ".csv"), 
          row.names = FALSE, )

write.csv(EHF_percentiles, 
          paste0(output_path, "/", 
                 deparse(substitute(EHF_percentiles)), 
                 "_2019_", 
                 nrow(EHF_percentiles), 
                 "stations", 
                 ".csv"), 
          row.names = FALSE, )

write.csv(ECF_percentiles, 
          paste0(output_path, "/", 
                 deparse(substitute(ECF_percentiles)), 
                 "_2019_", 
                 nrow(ECF_percentiles), 
                 "stations", 
                 ".csv"), 
          row.names = FALSE, )

write.csv(ECE, 
          paste0(output_path, "/", 
                 deparse(substitute(ECE)), 
                 "_2019_", 
                 nrow(ECE), 
                 "station_x_day" , 
                 ".csv"), 
          row.names = FALSE, )

write.csv(EHE, 
          paste0(output_path, "/", 
                 deparse(substitute(EHE)), 
                 "_2019_", 
                 nrow(EHE), 
                 "station_x_day" , 
                 ".csv"), 
          row.names = FALSE, )

write.csv(EHE_ECE, 
          paste0(output_path, "/", 
                 deparse(substitute(EHE_ECE)), 
                 "_2019_", 
                 nrow(EHE_ECE), 
                 "station_x_day" , 
                 ".csv"), 
          row.names = FALSE, )
```
